---
title: "Assignment 7 - Web scraping"
author: "Mie Arnau Martinez"
date: "10/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
# Loading relevant packages
library(pacman)
p_load(xml2)
p_load(rvest)
p_load(tidyverse) # Include tidyr, dplyr and stringr
p_load(janitor)
```

Webcraping of the worldometers website. Among others, this website holds information about population by country. 
```{r}
# Saving the URL to an object
url <- "https://www.worldometers.info/world-population/population-by-country/"
# scraping the website 
url_html <- read_html(url)

# Saving the data from the html tag "table" that captures the table from the website
table_list <- url_html %>% 
 html_nodes("table") %>%
 html_table()
# This is saved as a list

# Therefore, we unlist and bind together the separate lists as different columns of a dataframe.
table_df <- as.data.frame(do.call(cbind,unlist(table_list, recursive = FALSE))) 
head(table_df)

# Some of the column names are quite long so I rename the one's that I am going to use
table_df <- table_df %>% 
  rename(
    Country = "Country (or dependency)",
    Population = "Population (2020)",
    FertilityRate = "Fert. Rate",
    Density = "Density (P/Km²)"
    )

```


From this table, I want to visualize the fertility rate against the density (population / land area km²) to see if there is a relation between how dense the population is and how many children the women get on average.

```{r}
# Data cleaning
# After inspecting the data, I found that commas where used to delineate the numbers in the density and population column. Therefore, I start by removing these and transforming the values to numeric values
table_df$Population <- as.numeric(gsub("\\,", "", table_df$Population))
table_df$Density <- as.numeric(gsub("\\,", "", table_df$Density)) 
table_df$FertilityRate <- as.numeric(table_df$FertilityRate)

# Some of the values in FertilityRate are characters ("N.A."). Therefore, these cannot be made numeric and instead they are now labeled as NA.

# I then create some summaries of the data (mean population, mean density and mean fertility). To do this I first have to select the columns and filter out the NA's from FertilityRate.
table_df %>% 
  select(Population, Density, FertilityRate) %>% 
  filter(!is.na(FertilityRate)) %>% 
  summarise(mean_pop = mean(Population), 
            mean_dens = mean(Density), 
            mean_fertil = mean(FertilityRate))
```

From this, we can see that the average population per country is 38.8 million. Furthermore, the average population density is 359 people/Km² and the average female gets 2.7 children in their lifetime. 

Now to the visualization:
I want to plot the fertility rate against population density for the countries with a population greater than the average population. This is mainly to decrease the number of countries included in the plot so that it is more tangible. I also plot all countries at once to compare.
```{r}
# Expanding the scientific notation in the legend to whole numbers
options(scipen=999)

# Plotting the Density and FertilityRate for countries with a population above average
table_df %>%  
  filter(Population > 39000000) %>% 
  ggplot(aes(x = Density, y = FertilityRate, size = Population, color = Country)) +
    # Making the points more see-through by specifying the alpha < 1.
    geom_point(alpha=0.7) +
    # Changing the x axis scale to logarithmic to spread out the data points and pull outliers closer to the other data points
    scale_x_log10() +
    # Changing the scale for the size of the data points
    scale_size(range = c(1, 5)) + 
    # Specifying the range of the y axis
    ylim(0, 8) +
    # Specifying the title, x label and y label.
    labs(title = "Relation between Fertility Rate and Population Density in Greatly \n Populated Countries in 2020", 
         x = "Population Density", 
         y = "Fertility Rate") +
    # Specifying the theme
    theme_classic() + 
    # From here I just play around with the position and size of the different elements
    theme(legend.title = element_text(size = 7), 
          legend.text = element_text(size = 6),
          legend.position = "right",
          legend.direction = "horizontal") +
    # The last two lines decides the position of the title of the size and color legend and how many rows there should be for the color legend
    guides(color = guide_legend(title = "Color", title.position = "top", nrow = 9),
           size = guide_legend(title = "Population", title.position = "top"))


# Plot of the Density and FertilityRate for all countries
table_df %>%  
  ggplot(aes(x = Density, y = FertilityRate, size = Population, color = Country)) +
    geom_point(alpha = 0.7) +
    scale_size(range = c(1, 5)) + 
    scale_x_log10() +
    ylim(0, 8) +
    labs(title = "Relation between Fertility Rate and Population Density in 2020", 
         x = "Population Density", 
         y = "Fertility Rate")+
    # Color codes for each country couldn't fit in the plot so therefore they are excluded
    guides(color = "none")+
    theme_classic()

```

Of course, more statistical test are needed to conclude anything about the relationship between fertility rate and population density. However, from the plots there doesn't seem to be a trend in the data. This pattern is true both for the countries with the highest population as well as all countries in general.

References:
I used this guide for finding out how to re-position the elements in the plots: https://ggplot2.tidyverse.org/reference/guides.html
