---
title: "Web scraping and data preprocessing"
author: "Mie Arnau Martinez"
date: "12/20/2020"
output: html_document
---

This script provides the code for data acquisition and preprocessing for the project of making world map visualizations and investigating the effect of the Covid-19 crisis on the world economy.

A prerequisite for the current script is the csv file "raw_GDP_per_country.csv". This can be found on "https://github.com/Digital-Methods-HASS/au616353_martinez_mie/tree/master/my_final_project/data".
Similarly, the results can be found on the same Github folder.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries

```{r}
library(pacman)
 p_load(xml2)
 p_load(rvest)
 p_load(tidyverse) # Include tidyr, dplyr and stringr
```

Data acquisition of corona-related deaths per country. This was done through web scraping of Statista.

```{r corona related deaths data}
#------------------------
#   WEB SCRAPING
#------------------------ 
# Saving the URL to an object
url <- "https://www.statista.com/statistics/1093256/novel-coronavirus-2019ncov-deaths-worldwide-by-country/"
# scraping the website 
url_html <- read_html(url)

# Saving the data from the html tag "body" that captures the statistics from the website
corona_death <- url_html %>%
  rvest::html_nodes('body') %>%
  xml2::xml_find_all("//div[contains(@class, 'panelCard panelCard--statistic')]") %>%
  rvest::html_text()
# This is saved as a string so data cleaning is needed to make this into a data frame

#------------------------
#   DATA CLEANING
#------------------------ 
# Removing the title
corona_death <- str_replace(corona_death, "Number of deaths", "")
# Replacing "," with "."
corona_death <- str_replace_all(corona_death, ",", ".")
# Grouping the strings in countries and numbers of deaths and comma separate them
corona_death <- gsub("([^0-9]+)([^a-zA-Z]+)", "\\1,\\2,", corona_death)
# This allows us to only pick the countries
corona_country <- gsub("([^0-9]+)([^a-zA-Z]+)", "\\1", corona_death) 
# These are then saved as a list 
corona_list <- as.list(strsplit(corona_country, split = ",")[[1]]) 
# Inspecting the list, we see that element 198, 199 and 200 needs removal
corona_list <- corona_list[-c(198:200)]

# The same is done for the numbers
corona_number <- gsub("([^0-9]+)([^a-zA-Z]+)", "\\2", corona_death) 
corona_list2 <- as.list(strsplit(corona_number, split = ",")[[1]])
corona_list2 <- corona_list2[-c(198:199)]
# The last element contains "\n" so we replace it with only the correct value
corona_list2[[197]] <- '1'


# Merging the lists into one data frame
corona_list <- as.vector(unlist(corona_list))
corona_list2 <- as.vector(unlist(corona_list2))

corona_df <- as.data.frame(cbind("Country" = corona_list, "Deaths" = corona_list2))

# Lastly, I save it to a csv file
write_csv(corona_df, "../data/corona_deaths.csv")

```

Data preprocessing of GDP changes per country in 2019 and 2020.

```{r GDP changes per country data}
#------------------------
#   LOADING THE DATA
#------------------------
# Read csv file of the GDP changes from data folder
gdp_per_country <- read_csv('../data/raw_GDP_per_country.csv')
# Only selecting the relevant columns
gdp2019_2020 <- gdp_per_country[, 41:42]
# Adding the column with the countries
gdp2019_2020 <- cbind(gdp_per_country[,1], gdp2019_2020)

#------------------------
#   DATA CLEANING
#------------------------ 
# First, changing the string answers ("no data") to NA
gdp2019_2020$'2019'[gdp2019_2020$'2019' == "no data"] <- NA
gdp2019_2020$'2020'[gdp2019_2020$'2020' == "no data"] <- NA
# Removing all NA's
complete_gdp2019_2020 <- na.omit(gdp2019_2020)

# Removing regions and analytical groups
complete_gdp2019_2020 <- complete_gdp2019_2020[1:194,] %>% 
  rename(Country = "Real GDP growth (Annual percent change)")

# Lastly, I save it to a csv
write_csv(complete_gdp2019_2020, "../data/GDP_per_country.csv")
```

